Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-05-22 paper_jasa.tex
finished with intro, and have re-titled the following sections

However, measuring the strength of patterns seen in plots, and differences in individual perceptual ability, is something that is difficult and perhaps handicaps graphics  use among statisticians, where measuring probabilities is of primary importance. This has also been addressed in recent research. \citet{buja:2009} proposes a protocol that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between classical statistical inference procedures and exploratory data analysis. One of the protocols, the lineup, places the data plot among  a page of plots of null data, and asks a human judge to pick the plot that is different. Figure \ref{fig:test_category} shows an example lineup. Which plot do you think is the most different from the others? Wrapped in a process that mirrors classical inference, where there is an explicit, a priori, null hypothesis, picking the plot of the data from the null plots represents a rejection of that null hypothesis. The null hypothesis typically derives from the task at hand, or the type of plot being made. The alternative encompasses all possible antitheses, all types of patterns that might be detected in the data plot, accounting for all possible deviations from the null without the requirement to specify these ahead of time. The probability of rejection can be quantified, along with Type I, and Type II error, and power can be defined and estimated.

@@

The protocol has only been informally tested until now. This paper describes research where the lineup protocol is compared head to head with the relevant classical test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario. The intended use of lineups is where there is no existing test, and unlikely to be ever any numerical test. The thought is though, that the classical setting provides a benchmark for how well the lineup protocol works.

@@

In this paper we compare the lineup protocol to classical statistical inference, specifically applied to the regression framework. We also present results of a human-subject study assessing the performance of individuals on lineup plots \citep{buja:2009} for testing significance of regression parameters.

Section \ref{sec:visual_test} describes the basic ideas of visual inference, Section \ref{sec:regression} applies these ideas to the case of inference for regression analysis. In Section \ref{sec:simulation} we outline the setup for the human-subject study and  and present results.

\section{Visual Statistical Inference} \label{sec:visual_test} 
Visual tests are non-parametric tests of the form $H_0$: the data plot is visually not distinguishable from a null plot vs $H_1$: the data plot is visually distinguishable from a plot generated consistently with $H_0$.

@@

Theoretical power for the regression parameters is derived in the next section.
\section{Inference for a Regression Model} \label{sec:regression}

@@

\section{Simulation Experiment} \label{sec:simulation}

>>>>

However, measuring the strength of patterns seen in plots, and differences in individual perceptual ability, is something that is difficult and perhaps handicaps graphics  use among statisticians, where measuring probabilities is of primary importance. This has also been addressed in recent research. \citet{buja:2009} proposes a protocol that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between classical statistical inference procedures and exploratory data analysis. One of the protocols, the lineup, places the data plot among  a page of plots of null data, and asks a human judge to pick the plot that is different. Figure \ref{fig:test_category} shows an example lineup. Which plot do you think is the most different from the others? Wrapped in a process that mirrors classical inference, where there is an explicit, a priori, null hypothesis, picking the plot of the data from the null plots represents a rejection of that null hypothesis. The null hypothesis typically derives from the task at hand, or the type of plot being made. The alternative encompasses all possible antitheses, all types of patterns that might be detected in the data plot, accounting for all possible deviations from the null without the requirement to specify these ahead of time. The probability of rejection can be quantified, along with Type I, and Type II error, and $p$-value and power can be defined and estimated.

@@

The protocol has only been informally tested until now. In the work described in this paper, the lineup protocol is compared head to head with the relevant classical test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario for visual inference. The intended use of lineups is where there is no existing test, and unlikely to be ever any numerical test. The thought is though, that the classical setting provides a benchmark for how well the lineup protocol works under controlled conditions, and will provide some assurance that they will work in scenarios where there is no benchmark. Testing is done using a human-subjects experiment using Amazon's Mechanical Turk \citep{turk}, using simulation to provide controlled conditions for assessing lineups. The results are compared with those of the classical test.

@@

%In this paper we compare the lineup protocol to classical statistical inference, specifically applied to the regression framework. We also present results of a human-subject study assessing the performance of individuals on lineup plots \citep{buja:2009} for testing significance of regression parameters.

The paper is organized as follows. Section \ref{sec:visual_test} defines terms as used in visual inference, and describes how to estimate the important quantities from experimental data. Section \ref{sec:regression} focuses on the application of  visual inference to linear models.  Section \ref{sec:simulation} describes the results of the three simulation experiments conducted to compare the power of the lineup protocol with the equivalent uniformly most powerful (UMP) and an analysis of the resulting data. The last section describes the next steps in this research.

\section{Definitions and Explanations for Visual Statistical Inference} \label{sec:visual_test} 
Visual tests are non-parametric tests of the form $H_0$: the data plot is visually not distinguishable from a null plot vs $H_1$: the data plot is visually distinguishable from a plot generated consistently with $H_0$.

@@

Theoretical power for the regression parameters is derived in the next section.

\section{Application of Visual Inference in Linear Models} \label{sec:regression}

@@

\section{Results of Simulation Experiments} \label{sec:simulation}
